Categorical and gradient aspects of wordlikeness\thanks{Thanks to Gene Buckley, Charles Yang, and audiences at NELS 43 for helpful comments.}

### 2012-12-09 1.tex
Small edits to chapter 1

There is only one concise explanation for the [dʒ]-[tʃ] and [k]-[ɡ] alternations, namely that Russian speakers have internalized a process of voice assimilation.
Now, it is also the case that voicing in underlying representations in Russian is ``nondistinctive in all but the last member of a cluster of obstruents'' \citep[283]{A74}.
It is possible to deny, however, that this restriction on underlying hetero-voiced obstruent clusters is part of the grammatical knowledge of Russian speakers. 
For instance, \citeauthor{PE} write:

@@

\noindent
\citeauthor{PE} go on to label constraints on underlying representations as ``extensional'', and thus irrelevant to generative grammar, in contrast to ``intensional'' statements like phonological rules.
Regarding the facts about Russian obstruents discussed so far, it seems to be clear that the alternation facts are in some sense more privileged.
Were it the case that Russian speakers had not internalized a process of obstruent voicing assimilation, the only alternative explanation for the forms in (\ref{rovs}) is a massive system of phonologically conditioned suppletive allomorphy.
In contrast, were there no contraint on underlying obstruent voicing, however, it is not immediately obvious that \emph{anything} would be different. 
Presumably, this is what \citeauthor{PE} mean when they refer to constraints on underlying representations as ``computationally inert''.
There are also many cases where restrictions on underlying representations hold of underlying representations but not of derived forms.
For instance, hiatus is exceptionally rare in native Turkish roots; of the handful of examples in the Turkish Electronic Living Lexicon \citep{TELL}, many appear to be compounds (e.g., \emph{ısıalan} `endothermic', cf.~\emph{ısı} `heat', \emph{alan} `taker').
Yet, numerous phonological processes give rise to hiatus in derived environments \citep[e.g.,][]{Kabak2007b}.
\citet{Silverman2000} observes that operations like reduplication and truncation also tend to introduce violations of constraints on underlying representations.
However, there are many facts which strongly suggest that speakers internalize phonotactic constraints.

@@

In English, sequences of adjacent obstruents which do not also agree in voice (e.g., \emph{a}[b.s]\emph{inth}) are quite rare within a word, and therefore a hetero-voiced obstruent cluster is a clue to the presence of a word boundary in running speech.
Infants \citep[e.g.,][]{Mattys2001b} and \citep{McQueen1998b,Norris1997} are both thought to use this heuristic for word recognition in experimental settings.\footnote{
    There is reason to distinguish the computations involved in using this word segmentation heuristic from those implicated by other phonotactic behaviors.
    Whatever the locus of [bnɪk]'s ill-formedness, for example, no segmentation into multiple words or morphs renders it a well-formed sequence.
    See \S\ref{ss:bh} for further discussion.}

@@

However, \citet{Shibatani1973} shows quite convincingly that not all wordlikeness contrasts can be expressed as constraints on URs.
In German, for instance, obstruent voicing is contrastive, but neutralizes finally: e.g., [ɡʀaːt]-[ɡʀaːtə] `ridge(s)' vs.~[ɡʀaːt]-[ɡʀaːdə] `degree(s)'.
By hypothesis, the latter root ends in /d/, so the constraint against final voiced obstruents is specific to surface representations.
Yet, \citeauthor{Shibatani1973} claims, German speakers judge voiced obstruent-final nonce words to be ill-formed.\footnote{
    Voicing of final obstruents is usually lost in German loanword adaptation: e.g., English \emph{hot dog} becomes [hɑt dɔk] \citep[506]{Ussishkin2003}.}

@@

\citet{Hayes2004b} argues that phonotactic learning occurs before lexical or phonological acquisition, and is therefore must be independent of other types of grammatical knowledge.
This is not fully consistent with the available experimental evidence available, however.

@@

Crucially, all of this occurs \emph{before} there is any evidence for phonotactic knowledge, for which the earliest evidence begins at approximately 9 months of age \citep{Friederici1993,Jusczyk1993b,Jusczyk1994}.
While there are many gaps in current understanding (and keeping in mind the usual caveats about a strong interpretation of negative results), there is no reason to think that phonotactic knowledge is acquired before non-trivial amounts of lexical and phonological acquisition.
If anything, infants's phonotactic knowledge appears to come online at approximately the same point as highly specific phonological representations including subsyllabic units and the ability to extract allophonic generalizations, as predicted by the firm link between phonological and phonotactic knowledge advocated for here.

>>>>

There is only one concise explanation for the [dʒ]-[tʃ] and [k]-[ɡ] alternations, namely that Russian speakers have internalized a process of voice assimilation.
It is also the case that voicing in underlying representations in Russian is ``nondistinctive in all but the last member of a cluster of obstruents'' \citep[283]{A74}.
It is possible to deny, however, that this restriction on underlying hetero-voiced obstruent clusters is part of the grammatical knowledge of Russian speakers. 
For instance, \citeauthor{PE} write:

@@

\noindent
\citeauthor{PE} go on to label constraints on underlying representations as ``extensional'', and thus irrelevant to generative grammar, in contrast to ``intensional'' statements like phonological rules.
Regarding the facts about Russian obstruents discussed so far, it seems to be clear that the alternation facts are in some sense more privileged.
Were it the case that Russian speakers had not internalized a process of obstruent voicing assimilation, the only alternative explanation for the forms in (\ref{rovs}) is a massive system of phonologically conditioned suppletive allomorphy.
In contrast, were there no contraint on underlying obstruent voicing, however, it is not immediately obvious that \emph{anything} would be different. 
Presumably, this is what \citeauthor{PE} mean when they refer to constraints on underlying representations as ``computationally inert''.
There are also many cases where restrictions on underlying representations are specific to underlying representations and do not apply to surface representations.
For instance, hiatus is exceptionally rare in native Turkish roots; of the handful of examples in the Turkish Electronic Living Lexicon \citep{TELL}, many appear to be compounds (e.g., \emph{ısıalan} `endothermic', cf.~\emph{ısı} `heat', \emph{alan} `taker').
Yet numerous phonological processes give rise to hiatus in derived environments \citep[e.g.,][]{Kabak2007b}.
\citet{Silverman2000} observes that operations like reduplication and truncation also tend to introduce violations of constraints on underlying representations.
However, there are many facts which strongly suggest that speakers internalize phonotactic constraints.

@@

In English, sequences of adjacent obstruents which do not also agree in voice (e.g., \emph{a}[b.s]\emph{inth}) are quite rare within a word, and therefore a hetero-voiced obstruent cluster is a clue to the presence of a word boundary in running speech.
Infants \citep[e.g.,][]{Mattys2001b} and adults \citep{McQueen1998b,Norris1997} are both thought to use this heuristic for word recognition in experimental settings.\footnote{
    There is reason to distinguish the computations involved in using this word segmentation heuristic from those implicated by other phonotactic behaviors.
    Whatever the locus of [bnɪk]'s ill-formedness, for example, no segmentation into multiple words or morphs renders it a well-formed sequence.
    See \S\ref{ss:bh} for further discussion.}

@@

However, \citet{Shibatani1973} shows quite convincingly that not all wordlikeness contrasts can be expressed as constraints on URs.
In German, for instance, obstruent voicing is contrastive, but neutralizes finally: e.g., [ɡʀaːt]-[ɡʀaːtə] `ridge(s)' vs.~[ɡʀaːt]-[ɡʀaːdə] `degree(s)'.
By hypothesis, the latter root ends in /d/, so the constraint against final voiced obstruents is specific to surface representations.
\citeauthor{Shibatani1973} claims, however, that German speakers judge voiced obstruent-final nonce words to be ill-formed.\footnote{
    Voicing of final obstruents is usually lost in German loanword adaptation: e.g., English \emph{hot dog} becomes [hɑt dɔk] \citep[506]{Ussishkin2003}.}

@@

\citet{Hayes2004b} argues that phonotactic learning occurs before lexical or phonological acquisition, and it is therefore must be independent of other types of grammatical knowledge.
This is not fully consistent with the available experimental evidence, however.

@@

Crucially, all of this occurs \emph{before} there is any evidence for phonotactic knowledge, for which the earliest evidence begins at approximately 9 months of age \citep{Friederici1993,Jusczyk1993b,Jusczyk1994}.
While there are many gaps in current understanding (and keeping in mind the usual caveats about a strong interpretation of negative results), there is no reason to think that phonotactic knowledge is acquired before non-trivial amounts of lexical and phonological acquisition.
If anything, infants's phonotactic knowledge appears to come online at approximately the same point as highly specific phonological representations including subsyllabic units and the ability to extract allophonic generalizations, as predicted by the firm link between phonological and phonotactic knowledge argued for here.
